Cross-relaxation in dense ensembles of NV centers and application to magnetometry \\ \medskip Relaxation croisées d'ensembles de centres NV et application à la magnétométrie

### 2022-10-20 Manuscrit Thèse/These.tex
tata yoyo
toto

\documentclass[a4paper]{article}
\usepackage[french]{babel}
\usepackage[utf8]{inputenc}
\usepackage[]{amsmath}
\usepackage[]{braket} % \bra, \ket etc
\usepackage{graphicx}

\title{Le titre}

\begin{document}
\maketitle
\section{Section 1}
\subsection{Subsection 1}

\end{document}

>>>>

\documentclass[a4paper, 11pt]{report}

\usepackage{psl-cover}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric-comp,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{bib_thesis.bib} %Import the bibliography file

\usepackage{amssymb} %quelques symboles dont gtrsim /lesssim
\usepackage{subcaption} % package pour faire des subfigures
\usepackage{multirow} % package pour multirow/multicolumn
\usepackage{booktabs} % package pour top/mid/bottom rule
\usepackage{tcolorbox} % toujours plus de boites
\usepackage{xcolor} % Pour avoir des couleurs dans les équations

\title{Cross-relaxation in dense ensembles of NV centers and application to magnetometry \\ \medskip Relaxation croisées d'ensembles de centres NV et application à la magnétométrie}

\author{Clément Pellet-Mary}

\institute{ENS Paris}
\doctoralschool{EDPIF}{564}
\specialty{Physique}
\date{07 Décembre 2022}

\jurymember{1}{Dmitry Budker}{Professor JGU Mainz}{Rapporteur}
\jurymember{2}{Philippe Tamarat}{Professeur, Université de Bordeaux}{Rapporteur}
\jurymember{3}{Maria Chamarro}{Professeure, Sorbonne Université}{Examinateur}
\jurymember{4}{Vincent Jacques}{Directeur de recherche, CNRS}{Examinateur}
\jurymember{5}{Alexandre Tallaire}{Directeur de recherche, CNRS}{Examinateur}
\jurymember{6}{Gabriel Hétet}{Maître de conférences, ENS Paris}{Directeur de thèse}
% \jurymember{9}{Prénom NOM}{Titre, établissement}{Invité}
% \jurymember{10}{Prénom NOM}{Titre, établissement}{Invité}

\frabstract{
Le centre NV du diamant est un candidat prometteur pour de nombreuses technologies quantiques, que ce soit pour les communications quantiques, le calcul quantique ou la métrologie quantique. Grace aux progrès de synthétisation du diamant, des échantillons contenant une haute concentration en centres NV sont aujourd’hui réalisables. Cette augmentation de la densité de centres NV s’accompagne d’une augmentation des interactions entre ces derniers, ce qui donne lieu à une physique complexe et encore largement inexplorée. De nombreuses applications récentes, telles que l’hyper-polarisation en RMN, la bio-imagerie ou encore la spin-mécanique repose sur des ensembles denses de centres NV. Comprendre et maîtriser les propriétés d’ensemble de ces derniers est un point crucial pour l’établissement de ces technologies.

Cette thèse étudie les échanges de polarisation, par relaxation croisée, entre des ensembles denses de centres NV, ainsi que les interactions entre les centres NV et d’autres impuretés paramagnétiques présentes dans le diamant. Le but de ces travaux est de mieux comprendre les propriétés des ensembles de centres NV, et d’exploiter certaines de ces propriétés pour de nouvelles applications telles que la détection d’impuretés ou la magnétométrie.

Dans un premier temps, nous détaillerons les propriétés des centres NV et nous introduirons des concepts liés à l’interaction dipolaire entre spins et aux relaxations croisées. Nous verrons ensuite comment les relaxations croisées peuvent être utilisées pour détecter optiquement des résonances de spins qui ne sont pas optiquement actifs, en l’occurrence les défauts VH- et War1 dans des diamants CVD. Enfin nous étudierons les relaxations croisées entre différents sous-groupes de centre NV. Nous verrons d’abord leur influence sur la dynamique du spin et leur origine microscopique, puis nous étudierons leur dépendance avec le champ magnétique et leur potentielle utilisation pour la magnétométrie en champs faible.
}

\enabstract{
  The NV center in diamond is a promising candidate for emerging quantum technologies, including quantum communication, quantum computing and quantum metrology. Thanks to the recent progress in diamond synthesis, samples with high NV center concentration can now be created. This increase in the NV density comes with an increase of the interaction between them, which leads to a rich and complex physics. Numerous applications, such as NMR hyperpolarization, bio-imagery or spin mechanics rely on the use of dense NV ensembles. Understanding and mastering the properties of ensembles of NV centers will be a crucial point in the development of these technologies.

This doctoral thesis focuses on cross-relaxation between dense ensembles of NV centers, as well as on the interaction between NV centers and other diamond paramagnetic impurities. The aim of this work is to better understand the properties of dense NV centers ensembles, and to exploit them for new applications such as the detection of impurities and magnetometry.

In the first part, we will detail the properties of single NV centers and introduce notions regarding dipole-dipole spin coupling and cross-relaxations. We will then show how cross-relaxations can be used to optically detect dark spin resonances, in this case the VH- and War1 defects in CVD-grown diamond. Finally, we will present a study of cross-relaxations between subgroups of NV centers. We will first cover the influence of cross-relaxation on the spin dynamics and discuss its microscopic origin. We will then investigate the role of the magnetic field on cross-relaxations as well as their potential use in low field magnetometry.
}

\frkeywords{centre NV, dynamique de spin, magnétométrie}
\enkeywords{NV center, spin dynamics, magnetometry}

\begin{document}

\maketitle{}
\setcounter{tocdepth}{1}

\tableofcontents

\chapter*{Introduction}

Quantum mechanics was first developed at the beginning of the 20th century to explain physical observations such as the blackbody radiation \citep{planck1900theorie} or the photoelectric effect \citep{einstein1905erzeugung}. The quantum theory was subsequently developed in the first half of the 20th century and showed great success, in particular in giving a physical framework for chemistry \citep{pauling1931nature} and leading to a microscopic understanding of matter \citep{kittel1996introduction}. In the second half of the 20th century, technological development based on the new understanding of light and matter have led to what was retroactively called the first quantum revolution \citep{thew2019focus}. Technologies developed in the 50s and 60s include the semiconductor transistor \citep{bardeen1948transistor}, the laser \citep{maiman1960stimulated}, and nuclear magnetic resonance \citep{abragam1961principles} which have since revolutionized the fields of information science, global communications and medical imaging.

A new field of quantum technology has emerged in the last two decades. This ``second quantum revolution" differs from the first one in the usage of individual quantum systems, such as single photons \citep{gleyzes2007quantum}, atoms \citep{neuhauser1980localized} or electrons \citep{peil1999observing}, which have properties that can not be emulated by larger classical systems. These properties include quantum entanglement, superposition or measurement and form the basis of the newly formed quantum information science \citep{nielsen2002quantum, vedral2006introduction, hayashi2006quantum}.

Central to quantum information science is the idea of quantum bits or qubits \citep{schumacher1996sending}, which are the building blocks of quantum information in analogy to the bits of classical information theory. Physical implementation of a qubit could be any quantum system with two well defined quantum states, as long as these states can be initialized, manipulated and readout \citep{divincenzo2000physical}. Popular qubit candidates include superconducting circuits \citep{nakamura1999coherent, orlando1999superconducting}, photons \citep{bennett1992quantum, zhong2020quantum}, quantum dots \citep{veldhorst2014addressable, zajac2018resonantly}, trapped ions \citep{friis2018observation, wright2019benchmarking} and single crystal defects \citep{jelezko2006single, baranov2011silicon, zhong2015optically}.

Defects in large band gap crystals are sometimes referred to as artificial atoms \citep{buluta2011natural}, as the localized wave function of the electrons surrounding the impurity  give rise to discrete energy levels similar to that of an atom. Of all the possible crystals and defects, the most studied defect for quantum applications is the negatively charged nitrogen-vacancy center in diamond \citep{aharonovich2016solid, de2021materials}, often abbreviated to ``NV center". Although this defect had been known and correctly identified almost 50 years ago \citep{davies1976optical}, interest in the NV center really sparked after it was first isolated in 1997 \citep{gruber1997scanning}. The detection of single NV centers proved that NV centers fluoresce brightly \citep{gruber1997scanning}, are photostable \citep{kurtsiefer2000stable}, and that their electronic spin can be polarized and readout optically \citep{jelezko2004observation}, making them an promising solid state qubit.

Since then, single NV centers have been successfully used to create quantum entanglement over several km \citep{hensen2015loophole}, to store quantum information at room temperature for more than one second \citep{maurer2012room}, to detect single proteins via nuclear magnetic resonance \citep{lovchinsky2016nuclear}, and for many other applications in quantum communication \citep{wehner2018quantum}, quantum computing \citep{de2021materials} or quantum sensing \citep{degen2017quantum}.

As the interest in single NV center grew, diamond synthesis improved significantly \citep{achard2020chemical, barry2020sensitivity, edmonds2020generation}  which benefited not only single NV centers but also NV center ensemble. While single NV center are required for most quantum information processes \citep{ladd2010quantum}, several other applications benefit from a greater concentration of defects and are therefore more suited to operate with NV centers ensemble. Four of these applications are pictured in Fig. \ref{shema introduction}.

\begin{figure}[h!]
\centering
\includegraphics[width=.9\textwidth]{Intro et conclusion/Figures/Fig_intro}
\caption{Example of applications with dense ensemble of NV centers. (a) In vivo nanodiomond biomarkers (from \citep{mohan2010vivo}). (b) Quantum sensing with NV ensemble (from \citep{ku2020imaging}). (c) Dynamic nuclear polarization of an external spin bath (from \citep{healey2021polarization}). (d) Spin mechanics with trapped micro diamonds (from \citep{delord2020spin}).}
\label{shema introduction}
\end{figure}

The first application for NV center ensemble is the use of nano or micro diamonds as fluorescent biomarkers \citep{fu2007characterization, mohan2010vivo}. NV centers are good candidates for biomarkers as they emit a bright and stable photoluminesecence, and because diamond nanocrystals present low toxicity \citep{schirhagl2014nitrogen}. Some applications also use the spin degree of freedom of the NV centers, either to perform background free fluorescence microscopy \citep{chapman2013background} or to monitor the angular motion of the nanodiamonds \citep{mcguinness2011quantum, feng2021association}. For most of these applications, NV ensemble provide a better visibility than single NV centers and are therefore preferred.

The second NV ensemble application is for quantum sensing \citep{degen2017quantum}. NV centers are predominantly used as magnetic field sensors \citep{rondin2014magnetometry, barry2020sensitivity}, although they can also detect electric fields \citep{dolde2011electric, michl2019robust}, temperature \citep{acosta2010temperature, kucsko2013nanometre} ,crystal strain \citep{ovartchaiyapong2014dynamic, doherty2014electronic} or rotation \citep{ledbetter2012gyroscopes, ajoy2012stable}. Both single and ensemble of NV centers can be used for sensing: single centers offer unmatched spatial resolution \citep{mitchell2020colloquium} down to a few nm \citep{lovchinsky2016nuclear}, while ensemble provide better sensitivities \citep{wolf2015subpicotesla, barry2020sensitivity}.

The third NV ensemble application is nuclear magnetic resonance (NMR) dynamical nuclear polarization \citep{eills2022spin}. Dynamic nuclear polarization consists in increasing the nuclear polarization of a sample by transferring electron polarization to the nuclei bath \citep{abragam1978principles}. Because the NV electronic spin can be optically polarized at room temperature, micro and nano diamonds containing NV centers are promising candidates as hyperpolarization agents \citep{tetienne2021prospects}. Nuclear polarization of spins inside the diamond lattice was demonstrated with single NV centers \citep{jacques2009dynamic, smeltzer2009robust} and NV ensemble \citep{king2015room, scheuer2016optically, schwartz2018robust}. Polarization transfer to nuclear spins outside of the diamond is an ongoing challenge \citep{eills2022spin, healey2021polarization}, but dense NV ensemble are believed to be necessary to achieve high external polarization \citep{tetienne2021prospects, rizzato2022polarization}.

Finally, the fourth NV ensemble application is the recent field of spin mechanics \citep{perdriat2021spin}. Similarly to optomechanics \citep{aspelmeyer2014cavity}, the goal of spin mechanics is to couple a quantum system to the mechanical degree of freedom of a macroscopic object. Strong coupling between single NV center and an external mechanical oscillators has been observed \citep{rabl2009strong, kolkowitz2012coherent}, but coupling between NV centers and the motion of the diamond has so far only been achieved with dense NV ensemble \citep{delord2020spin, pellet2021magnetic, perdriat2022angle}. Coupling several NV centers to the same mechanical oscillator could also lead to interesting magnetic phase transitions \citep{wei2015magnetic, ma2017proposal}.

Most of the NV ensemble experiments discussed previously work under the assumption that the ensemble is constituted of $N$ independent single NV centers. However, as often in physics, \textit{more is different} \citep{anderson1972more}. Interaction between the NV center themselves or with their environment can lead to significantly different behavior between ensemble and single spins. These difference include a modification of the charge state dynamics \citep{giri2018coupled} and of the spin dynamics \citep{dobrovitski2008decoherence, jarmola2012temperature, mrozek2015longitudinal, choi2017depolarization}, as well as the apparition of exotic phenomena such as time crystal phase \citep{choi2017observation}, Anderson localization \citep{kucsko2018critical} or superradiance \citep{bradac2017room, angerer2018superradiant}.

The work presented in this manuscript focuses on dipolarly coupled dense ensemble of NV centers. More specifically, we look at spin polarization exchange between the NV centers themselves and with their environment. Polarization exchange (flip-flop, spin diffusion, cross-relaxation, ...) in dense ensemble of spins has been abundantly studied in the context of NMR \citep{abragam1978principles} and spintronics \citep{vzutic2004spintronics}. Because NV centers tend to be relatively dilute, the question of spin exchange between between NV themselves \citep{choi2017depolarization} or with their environment \citep{hall2016detection} has only recently been addressed. However, as samples with higher NV concentration become available \citep{acosta2009diamonds, tallaire2020high, shenderova2019synthesis}, these aspects are becoming more and more relevant.

The goal of this present work is both to better understand the physics of dense NV ensemble, which is crucial for many of the NV ensemble applications cited previously, and to exploit some of these many-body properties for new applications.

\bigskip
This manuscript is organized in the following way:

\medskip
The first chapter introduces the theoretical and experimental concepts used in this manuscript. This includes a brief overview of diamond and NV centers fabrication processes, the physics of single NV centers and a presentation of magnetic dipole-dipole interaction and cross-relaxation.

The second chapter covers the observation of cross-relaxation between NV centers and other spin impurities in CVD-grown diamond. We use measurements on the NV centers to identify these impurities and to quantify some of their attributes. Most of the results of this chapter have been published in  \citep{pellet2021optical} and \citep{ngambou2022improving}.

The third chapter focuses on the dipole-dipole mediated spin relaxation within dense ensemble of NV centers. Following the work presented in \citep{choi2017depolarization}, we introduce the concept of NV-flutcuator to explain this phenomenon and present experimental and theoretical results solidifying this hypothesis. The results of this chapter have been partly published in \citep{pellet2022spin} and \citep{pellet2021magnetic}.

Finally, the fourth chapter presents a study on NV-NV dipolar relaxation under low magnetic field. This chapter covers both the rich physics of NV centers under low magnetic field, and a potential application of the low field depolarization as a new magnetometry protocol, as well as comparisons with other NV magnetometry protocols. The results of this chapter have in large part been published in \citep{pellet2022spin}.

\chapter{Spin properties of single and ensemble NV centers}
The aim of this first chapter is to introduce concepts related to Nitrogen-vacancy (NV) centers, to the diamond material, and to the dipole-dipole interaction which are necessary to explain the work presented in the later chapters.

In this introductory chapter, we will first cover the material aspects of NV centers by focusing on the creation process of synthetic diamonds and the required steps to create NV defects inside them. We will then cover the fundamentals of NV physics, including the NV center optical and spin properties, basic experiments with NV centers and the various relaxations times ($T_1$,$T_2$,$T_2^*$) which characterizes the NV center's spin. Finally we will present the dipole-dipole interaction between two spins and the concept of cross-relaxation.

\section{The NV center as a physical object}
\begin{figure}[h!]
\centering
\includegraphics[width=0.9\textwidth]{chapter 1/Figures/diamant_et_NV}
\caption{From the diamond to the NV center. a) Drawing of a bulk diamond\footnote{In this manuscript, we will consider any samples above $\sim$ 100 $\mu$m to be bulk.}. b) Drawing representing a fluorescence microscopy image of a diamond surface. The red fluorescent spots symbolize NV centers photoluminescence. c) Zoom on the crystalline structure of an NV center. Black balls represent carbon atoms}
\label{diamond+NV}
\end{figure}
Before studying the NV centers properties, we must first discuss the diamond itself and answer these questions: How are diamond created, what kind of defects are there in diamond, and how to specifically create NV defects. The field of material science related to diamonds and NV centers is a rich and complex one, and this section will only touch upon basic notions regarding these subjects.

Fig. \ref{diamond+NV} illustrates the various scales to consider in the study of NV centers. Starting with a bulk diamond, typically few mm wide, then zooming-in on the diamond surface with an optical microscope to find fluorescent spots corresponding to the photoluminescence of NV centers. Standard confocal microscopy is limited in its resolution to the $\sim \mu$m range, but if it was possible to zoom-in even further, we would see the atomic structure of a few \AA \ illustrated in the last panel. 
\subsection{Diamond material overview}
\begin{figure}[h!]
\centering
\includegraphics[width=0.6\textwidth]{chapter 1/Figures/phase_diagram_carbon_wiki}
\caption{Pressure-temperature phase diagram of elemental carbon. Credits: Wikimedia commons, adapted from \citep{bundy1989pressure, bundy1996pressure}.}
\label{carbon phase diagram}
\end{figure}

Diamond is a remarkable material in many respects. It is famous for being the hardest and most thermally conductive non-synthetic material. It is also a material with high optical refractive index and optical dispersion which makes it an extremely valuable crystal in jewelry.

Diamond is also famous for being extremely rare in nature, even though its only constituent, carbon, is one of the most abundant element on the Earth surface. The reason being that graphite, and not diamond, is the thermodynamically stable phase of carbon under ambient conditions.

Fig. \ref{carbon phase diagram} shows the PT phase diagram of elemental carbon. We can see that diamond becomes the thermodynamically stable phase only for pressure above a few GPa, $10^4$ times higher than atmospheric pressure. Even though diamond is unstable under ambient conditions, the extremely high energy activation needed to break the sp$^3$ bond between two carbon atoms means that the kinematics of the $C_{\rm diamond} \to C_{\rm graphite}$ transition is almost completely frozen. In many ways diamonds are more stable than graphite under ambient condition due to their relative chemical inertness.

Where the thermodynamic equilibrium really matters is for the crystallization process. The higher pressure needed to reach the diamond-stable region can only be found naturally deep under the earth crust, typically between 150 and 240 km below sea level \citep{tappert2011diamonds}, which explain their scarcity. It also explains why, despite many previous attempts, diamonds were only synthetically produced in 1953 \citep{barnard2000diamond}, more than 50 years after the first synthetic sapphires. To this day, artificial diamond synthesis remains an active field of research \citep{shenderova2019synthesis, achard2020chemical}.

Diamonds grown naturally or artificially are used in the lab and in industry for their extreme hardness, with applications ranging from abrasive to diamond anvil cells \citep{jayaraman1983diamond}. More recently, diamonds have been the object of optics and electronics applications. For both of these fields however, the focus is no longer solely on the diamond but also on the defects hosted inside it.

\subsection{Point defects in diamond}
\begin{figure}[h!]
\centering
\includegraphics[width=0.9\textwidth]{chapter 1/Figures/Diamond_type}
\caption{Traditional classification of diamonds, based on \citep{tappert2011diamonds}.}
\label{diamond type}
\end{figure}
Diamond in its intrinsic form is an insulator with a bandgap of $\sim 5.47\ \rm eV$ and is transparent from far infrared to deep UV. As any crystal however, diamond is prone to defects. We will focus here on point-like (0D) defects and omit structural defects such as dislocations, even though those too can affect the optical properties of the diamond \citep{collins2000colour}.

Point-like defects can be constituted by the absence of a carbon atom (``vacancies"), an interstitial defect between two carbon atoms or the substitution of a carbon atom with another atom. These impurities result in unpaired electron or holes localized around the impurity site. This localization causes discrete energy levels for the holes or the electrons, some of which lie inside the diamond bandgap and impact the crystal optical and electrical properties.

Defects with electronic transitions in the optical range are called ``colored centers" as they are responsible for the coloring of the gems.  Over 20 elements are known to form colored centers when introduced in diamond \citep{zaitsev2013optical, shenderova2019synthesis}, and each of these elemental atoms can form several defects. Nitrogen alone can form more than 50 different colored centers \citep{dobrinets2016hpht, ashfold2020nitrogen}.

Nitrogen, and to a lesser extant boron, are the most commonly found extrinsic elements in natural diamond. The traditional classification of diamond is presented in Fig. \ref{diamond type}. It is based on the N and B concentration, with a threshold of a few ppm (part per million) for each species as this was the smaller amount easily detectable through IR absorption. Type Ia diamonds contain clustered N defects such as the B-center (N$_4$V defect). In contrast, type Ib diamond mostly contains C-center which are single substitutional N$_s$ defects, also referred to as P1 centers in the spin community. Type IIb diamond contains boron impurities which give them a blueish-grey color, and type IIa diamond contains no detectable impurities via IR absorption \citep{ashfold2020nitrogen}.

Nitrogen-vacancy centers are a rare occurrence in both natural diamonds and untreated synthetic diamonds, but we will see later that the much more common N$_s$ defects can be converted in NV centers. When working with NV centers, the starting crystal is therefore often a type Ib diamond if one wants to work with ensemble, or a type IIa diamond if one wants to work with single NV centers \footnote{In order to isolate single NV in a confocal optical spot, the concentration needs to be [NV] $<$ 1 ppb.}.

The boron defects are mainly studied in the context of p-doped diamond semi-conductor: a concentration of [B] $\approx 1000\ \rm ppm$ is needed to achieve a significant overlap between the impurity sites, which turns the diamond into a semi-metal \citep{macpherson2015practical}. Other defects of interest, generally not present naturally but introduced voluntarily, include P and S defects to create n-doped diamond \citep{das2022diamond} (still an active field of research), as well as group IV-vacancy defects (SiV, GeV, SnV, PbV) for quantum optics application \citep{bradac2019quantum}.

The NV center however remains by far the most studied defect in diamond due to its unique spin properties which will be detailed below.

\subsection{Creation of synthetic diamond for optics and electronics applications}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/HPHT_and_CVD}
\caption{Schematics of HPHT (a) and CVD (b) growth process as detailed in main text}
\label{HPHT and CVD}
\end{figure}

Almost all diamonds used in optics or electronics experiments today are synthetic, due to the higher control of the growth process required for these applications. There are two main competing technologies for the growth of large ($\mu \rm{m} \sim \rm{mm}$) single crystal diamonds: High pressure high temperature growth (HPHT) and chemical vapor deposition growth (CVD). Both of these processes have their own specificities which are detailed below.

\subsubsection{HPHT growth}

HPHT was the first commercially viable process for synthetic diamonds, starting from the 1950s \citep{barnard2000diamond, bundy1955man}. The basic idea of HPHT synthesis is to reconstruct an environment where diamond is the thermodynamically stable phase of carbon, similarly to the growth process of natural diamond below the earth surface.

Fig. \ref{HPHT and CVD}-a) illustrates the apparatus of HPHT synthesis: a carbon source (graphite) and a solvent metal composed of transition metals (Fe, NI, Co typically \citep{bundy1963direct}) are put under $5 \sim 6$ GPa pressure in a press and heated to temperatures $\geq 1300\ ^\circ$C. A diamond crystal seed is put in the metal melt and a temperature gradient insures that the carbon migrates from the graphite to the diamond.

The first synthesis of HPHT diamonds were all of type Ib due to unwanted nitrogen pollution in the metal melt. The first HPHT type IIa diamonds were produced in the late 90s by adding other elements to the metal bath (Ti, Al, Zr) which would react preferentially with nitrogen and prevent its incorporation in the diamond \citep{burns1999growth, sumiya2002growth}.

\subsubsection{CVD growth}
\label{fab HPHT}

Diamond growth by CVD also started in the 50s, but only really took off in the 80s \citep{matsumoto1982growth, matsumoto1982vapor, kamo1983diamond} with the introduction of atomic hydrogen catalysis. Unlike HPHT, CVD operates in a regime where diamond is not the thermodynamically stable phase. It relies therefore heavily on catalysts to stabilize the diamond phase versus the graphite one.

The main catalyst used is atomic hydrogen because it etches the $C-C$ sp$^2$ bonds (graphite) faster than it does sp$^3$ bonds (diamond) \citep{gracio2010diamond}. In modern CVD growth, hydrogen is by far the most abundant element in the gas phase: the precursor gases used for the growth are typically CH$_4$ and H$_2$ in proportions of $1-5\%$ and $95-99\%$ respectively \citep{achard2020chemical}.

Fig. \ref{HPHT and CVD}-b) illustrates the apparatus of a microwave plasma reactor: it consists in a modified vacuum chamber which acts as a microwave resonator. The microwave creates a plasma ball from the the precursor gases with a temperature $T_{\rm plasma} \geq 3000 ^\circ$C \citep{ashfold2020nitrogen}, enough to dissociate the $H_2$ gas in atomic hydrogen \citep{balmer2009chemical}. The plasma acts as the carbon source for the diamond growing on top of the substrate, and as the catalyst to favor the diamond phase. %The substrate is kept at a temperature $700 \sim 1100\ ^\circ$C during the growth.

Other sources can be used to create the high temperature required for atomic hydrogen, such as a tungsten hot filament \citep{haubner1993diamond}, electrical discharge (arcjet) \citep{luque1998excited} or even an oxygen-acetylene combustion flame \citep{bachmann1991towards}.

If the CVD reactor is well built, the main sources of extrinsic impurities in the diamond comes from the source gasses \citep{balmer2009chemical}. The commercial availability of source gases containing less than $\sim$ ppm impurities made possible the growth of  high purity type IIa CVD diamonds \citep{kasu2003high, twitchen2004high, tallaire2006characterisation}.

\medskip
Both HPHT and CVD processes can produce diamonds of high quality for quantum optics or electronics application, each with their own pros and cons. HPHT offers better scalability for thick crystals ($\sim \rm mm$) and allows the incorporation of heavy atoms (Ge \citep{palyanov2015germanium} , Sn \citep{ekimov2019effect}) which can be hard to incorporate from a gas phase. CVD on the other hand allows more versatility in the control of the impurities introduced in the crystal, including an isotopic control on both the impurities and the carbon itself. A particular advantage of CVD growth is the possibility of $\delta$-doping where a specific defect is grown only within a thin layer of the crystal, as low as 5 nm \citep{ohno2012engineering, ishikawa2012optical, ohashi2013negatively}.

\subsection{Optimizing NV center formation}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/formtion_NV_2}
\caption{Illustration of the formation of NV centers. Black and green dots represents carbon and nitrogen atoms respectively}
\label{formation NV}
\end{figure}

NV centers are a relatively rare impurity in untreated diamonds. In CVD diamond for instance, the proportion of nitrogen atoms forming an NV center in an untreated crystal is at best of $2-3 \%$ \citep{hartland2014study}. While this ratio can be enough  to work with single NV centers, applications that require dense ensemble need a higher conversion rate. We will discuss here the usual techniques employed to improve the formation of NV centers.

Fig. \ref{formation NV} illustrates the steps needed to maximize the number of NV centers in the crystal while keeping a relatively high crystalline purity.

The first step to create NV centers is to get individual substitutional N atoms in the crystal ($N_s$, C-centers or P1 centers depending on the literature). This can be achieved either \textit{in situ} by introducing (voluntarily or not) nitrogen in the crystal during the growth process \citep{tallaire2006characterisation, lobaev2017influence}, or \textit{ex situ} by implanting N$^+$ ions with ion implantation techniques \citep{meijer2005generation, smith2019colour}. Because this present work focuses on dense NV ensembles, we will consider the simplest case where the starting crystal is a type-Ib diamond because of the presence of nitrogen during the growth. A starting concentration [N$_s$]= $10 \sim 300$ ppm is typical for type Ib HPHT diamonds \citep{achard2020chemical}.

The second step is to create monovacancies in the crystal lattice. This step might be optional for crystals which already contain a large number of monovacancies, but CVD crystals in particular can have less vacancies that they have substitutional nitrogen atoms \citep{mainwood1999point}. The general technique to create monovacancy is to bombard the sample with high energy particle in order to knock carbon atoms out of the crystal. Several high energy particles have been used to create vacancies, such as protons, electrons, ions, neutrons and even gamma rays \citep{davies1976optical, ashbaugh1988gemstone, kleinsasser2016high}. The most common irradiation source for for bulk diamond however is electron irradiation (e-beam) \citep{acosta2009diamonds}, thanks to their higher availability (compared to neutrons or gamma rays) and their ability to deeply penetrate the crystal.

The final step is to move the vacancies next to the nitrogen atoms. The activation energy required to move vacancy is of $\sim 2.3\ \rm eV$ which corresponds to temperatures of $600\sim 700 ^\circ$C \citep{davies1992vacancy, newton2002recombination}. The most common method to move the vacancies is therefore to anneal the crystal at a temperature $800 \sim 1200 ^\circ$C for a few hours \citep{botsoa2011optimal}. When the vacancy is next to a nitrogen atom, it will only become mobile again for temperatures of $1400 \sim 1500 ^\circ$C \citep{zaitsev2013optical, pinto2012diffusion} which explains why the annealing process favors NV creation.

The final N$_s$ to NV conversion rate can be as high as $\sim 50 \%$ \citep{grezes2015storage, hartland2014study}, although this ratio tends to decrease for higher [N$_s$] concentrations. Most sample studied in this manuscript had a concentration $[\rm NV]= 1-10\ \rm ppm$ (see sample details in Appendix \ref{Appendix samples})

%\subsection{Controlling the NV center charge state}
%
%The charge state of the NV center is of crucial importance for NV applications. Only the negatively charged NV$^-$ center shows the specific spin properties that will be detailed below, but the NV center can exist under three charge states in the diamond: NV$^-$, NV$^0$ and NV$^+$. The presence of NV$^+$ is usually marginal \citep{hauf2014addressing, pfender2017protecting} and we will focus here solely on the NV$^-$ and NV$^0$ charge state.
%
%The charge state of the NV centers depends not only on the growth process of the crystal, but also on the experiment performed with the sample. In particular, illumination in the optical range can promote electrons in the conduction band or holes in the valence band and modify the charge state of the NV center.
%
%On the material side, the crystal has to remain electrically neutral which means that at least as many charge donors need to be present as there are NV$^-$ centers. For nitrogen-rich samples, the donors are mostly N$^+$ centers, which imposes a limit on the maximum conversion rate of N to NV in order to keep a high [NV$^-$]/[NV$^0$] ratio. The possibility to introduce other donors, such as phosphorus, has been studied \citep{doi2016pure} but remains impractical so far \citep{barry2020sensitivity}.
%
%On the experimental side, it has been shown that the ionization of NV$^-$ to NV$^0$ and the reverse process depend on the intensity and wavelength of the laser used for illumination \citep{aslam2013photo}. It was found that the optimal wavelength to preserve the NV$^-$ charge state was  $510-540\ \rm nm$. The most common wavelength used with NV centers is 532 nm (green) due to the historical importance of Nd:YAG lasers. On the other hand, it is possible to semi-permanently ionize the NV$^-$ into NV$^0$ when using a red excitation, which has been used to create optical memories \citep{dhomkar2016long}.
%
%For samples in high excess of [N$_s$] such as type Ib HPHT diamonds, the question of the charge state of the NV center is of relatively low importance as the NV$^-$ state is the predominant one. For sample with a lower [N$_s$] content and a higher [NV]/[N$_s$] ratio however, one must be more careful with the optical intensity and wavelength used as the concentration of NV$^0$ can reach up to 50\% of the total NV concentration \citep{grezes2015storage}

\section{The NV$^-$ center optical and spin properties}

This section focuses on the NV$^-$ center properties both as an optical defect (colored center) and as a spin defect, and compare them to other solid state systems. It also covers the interplay between the spin and the optical levels, a property almost unique to the NV$^-$ center which is the basis for optical polarization and readout of the NV$^-$ spin state.

\subsection{NV center molecular orbitals}

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{chapter 1/Figures/NV_electrons}
\caption{(a) Representation of the nitrogen and carbon orbitals surrounding the vacancy of an NV center. Each black dot represents an electron. A sixth electron is required to form the negatively charged NV$^-$ state. (b) Molecular orbitals of the NV center predicted from the the $C_{3v}$ symmetry of the defect. The electron occupation of the NV$^-$ ground state is represented by red arrows.}
\label{NV electrons}
\end{figure}

We will first cover the molecular aspect of the NV defect in diamond and the energy levels of the defect predicted from group theory and ab initio computations.

Fig. \ref{NV electrons}-a) illustrates the $sp^3$ orbitals of the carbons and nitrogen atoms neighboring the vacancy, as well as the 5 electrons occupying dangling bonds from these atoms and a sixth electron captured to form the negatively charged NV$^-$ state.

Fig. \ref{NV electrons}-b) shows the predicted molecular orbitals of the NV centers, initially computed as linear combination of the $sp^3$ orbitals respecting the $C_{3v}$ symmetry of the defect \citep{loubser1978electron}. This model has since been confirmed by ab initio computation and experimental observations \citep{doherty2013nitrogen}. In particular ab initio computations have shown that the molecular orbitals of the NV center are mostly localized on the nearest neighbors of the vacancy \citep{gali2008ab}. Ab initio computation has also determined that the $a'_1$ level lies in the valence band and is thus often ignored.

The electron occupations shown in Fig. \ref{NV electrons}-b) corresponds to the electronic ground state of the NV$^-$ charge state.

In the remaining part of this manuscript, we will only consider the negatively charged NV$^-$ state, unless specified otherwise. We will therefore omit the ``$-$" from now on and simply refer to the negatively charged state as ``NV".
\subsection{Optical properties of the NV center}
\begin{figure}[h!]
\centering
\includegraphics[width=1\textwidth]{chapter 1/Figures/NV_optical}
\caption{Optical properties of the NV center. (a) Electron occupation of the molecular orbitals for the electronic ground state $^3A_2$ and the first excited state $^3E$. (b) Diagram of the main optical transition between the two spin triplet states and of the IR transition between the two singlet sates. The value of the transition corresponds to the zero phonon line (ZPL) and are taken from \citep{doherty2013nitrogen}. (c) Representation of the vibronic structure for the $^3A_2$ and $E^3$ states under the Franck-Condon approximation. The graph axes are the energy of the states $E$ and configuration coordinate $q$. The green and red arrow represents non-resonant (Stokes) absorption and emission processes. (d) Emission and absorption spectrum of the NV center main transition at 300 K, from PL and PLE measurements respectively. Credits: Wikimedia commons}
\label{NV optical}
\end{figure}

Because the NV center has discrete electronic transitions inside the diamond band gap, it fluoresces: it can absorb light from $450-650$ nm and re-emit light in the $600-800$ nm band.

Fig. \ref{NV optical}-a) shows the electronic ground state $^3A_2$ and first excited state $^3E$, where 1 electron from the $a_1$ molecular orbital is promoted to the $e$ level. Both of these electron occupations have two unpaired electrons, meaning that they can exist either either in a spin triplet state ($S=1$) or a spin singlet state ($S=0$) \footnote{The two singlet states $^1E$ and $^1A_1$ represented in Fig. \ref{NV optical}-b) are both of configuration $a_1^2e^2$ where the two $e$ electrons are either in a the same $e$ orbital, or in an antisymmetric superposition of the two $e$ orbitals \citep{doherty2011negatively}.}.
